---
title: "Classification"
author: "Nwadiugwu-A2-ISQA 8080"
date: "September 22, 2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Classification - Logistic Regression

Essential Classification algorithms using R. Implementing and evaluating Logistic Regression with R's glm function. 

```{r LogReg}

# set your working directory to the current file directory 
tryCatch({
  setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
  }, error=function(cond){message(paste("cannot change working directory"))
})

library(ISLR)
library(GGally)
library(Hmisc)
library(e1071)
library(pROC)
library(tidyverse)
library(caret)
library(skimr)
library(tidymodels)

# step 1:reading the file 
census <- read_csv("Census.csv")

```


```{r LogReg}

# the data preparation steps is performed in the Utils.R code 
source("Utils.R")
dataset <- prepare_census(census)
names(dataset)
dim(dataset)

#An overview of the data 
glimpse(dataset)

# nicer:
skim(dataset)

# Scatterplots and pairwise comparisons
ggpairs(census, aes(color = Income))

#Calculating correlation between the variables made easier by getting the numerical columns only
dataset_numeric <- dataset %>% select_if(is.numeric)
cor(dataset_numeric)

#Identifying correlated predictors using the caret package
(high_correlation <- findCorrelation(cor(dataset_numeric), cutoff = 0.7))


##### Logistic Regression #####

#specifying the model
target_var <- 'Income'
model_form <- Income ~ Age + Sex + Race + MaritalStatus + WeeklyHours
model_type <- "glm" # general linear model, which includes logistic regression
positive_class <- "Yes"
negative_class <- "No"

# we create a data partition (training-test) by specifying our target variable
set.seed(3456)
train_index <- createDataPartition(y = dataset[[target_var]], p = 0.7, list = FALSE) # creata a 70-30 split

Train <- dataset %>% slice(train_index)
Test <- dataset %>% slice(-train_index)
summary(Train)
summary(Test)


#Spliting data with Equal percentage split 
#set.seed(3456)
#trainIndex <- createDataPartition(dataset$Income, p = .7, list = FALSE, times = 1)
#Train <- dataset[ trainIndex,]
#Test <- dataset[-trainIndex,]
#summary(Train)
#summary(Test)

model_recipe <- recipe(model_form, data = Train, family='binomial') %>%  step_zv(all_predictors())


# step 3: specify training parameters and training the set

# resampling with a CV of 10. we add savePredictions and classProbs to get the probability estimates from the predictions
trControl <- trainControl(method='cv', number = 10, savePredictions = TRUE, classProbs = TRUE)

#Running the glm function on the train result
glm_fit <- train(model_recipe, data = Train, method = model_type, trControl = trControl)

# step 4: evaluate the model
# calling the summary function on the lm object will give you the basic regression output
summary(glm_fit)

coef(glm_fit$finalModel)

summary(glm_fit$finalModel)$coef
summary(glm_fit$finalModel)$coef[,1] #this gives you the point estimators only


# step 5: get the predictions and evaluate the model performance
# now, let's get the probabilities of Income (Yes or No) with the model
glm_probs <- glm_fit %>% predict(newdata = Train, type = "prob") # This gives predictions on the training set!

# show the first 10 predictions. 
glm_probs[1:10, 2] # the '2' indicates the probability of predicting 'Yes'. 

# Converting the above to binary predictions with a default threshold value of 0.5
threshold <- 0.3220074
glm_pred <- factor(ifelse(glm_probs[, positive_class] > threshold, positive_class, negative_class) , levels = levels(Train[[target_var]]))

# note: for the standard threshold, we can also use the predict() function directly
glm_pred_direct <- glm_fit %>% predict(newdata=Train, type='raw')

# we can get the evaluation metrics confusion matrix from the model output:
postResample(glm_pred, Train[[target_var]])

#Displaying the confusion matrix and additional metrics.
confusionMatrix(glm_pred, Train[[target_var]], positive = positive_class)

# note: the previous data prep steps and code set the class of interest to 'yes' and calculates sensitivity and specificity accordingly. we can change that by manually providing the 'positive' class
confusionMatrix(glm_pred, Train[[target_var]], positive = "No")

#looking at the ROC curve
(glm_roc <- roc(Train[[target_var]], glm_probs[, positive_class], plot = TRUE, print.auc = TRUE, legacy.axes = TRUE, levels = c(negative_class, positive_class)))

# get auc
(Train_auc <- auc(glm_roc))

# side note: how do we decide on the threshold to use?
# one option is: select the threshold that maximizes the combined sensitivity and specificity (other options available, see documentation)
(threshold <- coords(glm_roc, 'best')$threshold)


# we can also create a confusion matrix when we have a threshold value other than 0.5
#threshold <- 0.7
#glm_pred <- factor(ifelse(glm_probs[, positive_class] > threshold, positive_class, negative_class) , levels = levels(Train[[target_var]]))
# create the confusion matrix. note: the positive = 'Yes' parameter indicates that we see the 'Yes' outcome as the outcome of interest
#confusionMatrix(glm_pred, Train[[target_var]], positive = positive_class)
# finally, predict a specific value
#newdata <- data.frame(Age=c(20), Sex = factor(c("0")), RestBP = c(140), Chol = c(0.2),  Ca = factor(c("0")))

#Prediction using the Test set

tglm_pred = glm_fit %>% predict(newdata = Test, type="raw") #creates the predicted class

tglm_probs <- glm_fit %>% predict(newdata = Test, type = "prob") #This gives predictions on the test set!

# show the first 10 predictions. 
tglm_probs[1:10, 2] # the '2' indicates the probability of predicting 'Yes'.

# then, let's convert this to binary predictions with estimated "best" threshold value 
threshold <-  0.1702284 # from the threshold that maximizes the combined sensitivity and specificity
tglm_pred <- factor(ifelse(tglm_probs[, positive_class] > threshold, positive_class, negative_class) , levels = levels(Test[[target_var]]))

#The evaluation metrics and confusion matrix from the model output:
postResample(tglm_pred, Test[[target_var]])

# Displaying the confusion matrix and additional metrics.
confusionMatrix(tglm_pred, Test[[target_var]], positive = positive_class)

# note: the previous data prep steps and code set the class of interest to 'yes' and calculates sensitivity and specificity accordingly. we can change that by manually providing the 'positive' class
confusionMatrix(tglm_pred, Test[[target_var]], positive = "No")

#looking at the ROC curve

(glm_roc <- roc(Test[[target_var]], tglm_probs[, positive_class], plot = TRUE, print.auc = TRUE, legacy.axes = TRUE, levels = c(negative_class, positive_class)))

#Getting the auc
(Test_auc <- auc(glm_roc))

# side note: how do we decide on the threshold to use?
# one option is: select the threshold that maximizes the combined sensitivity and specificity (other options available, see documentation)
(threshold <- coords(glm_roc, 'best')$threshold)

```

## Linear Discriminant Analysis (LDA)

Now, let's look at LDA as an alternative model for classification. The MASS library has a ready-made lda() function that we'll use for this

```{r lda}

#using the preparation function for the dataset provided in the Utils.R script

dataset <- NULL
census <- read_csv("Census.csv")
source("Utils.R") # load the functions provided in Utils.R
dataset <- prepare_census(census)

# we create a data partition (training-test) by specifying our target variable
set.seed(3456)
train_index <- createDataPartition(y = dataset[[target_var]], p = 0.7, list = FALSE) # creata a 70-30 split

Train <- dataset %>% slice(train_index)
Test <- dataset %>% slice(-train_index)

# specifying the model
target_var <- 'Income'
model_form <- Income ~ Age + Sex + Race + MaritalStatus + WeeklyHours
model_type <- 'lda'
positive_class <- 'Yes'
negative_class <- 'No'

model_recipe <- recipe(model_form, data = dataset) %>% 
  step_novel(all_predictors(), -all_numeric()) %>%
  step_unknown(all_predictors(), -all_numeric()) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>% 
  step_zv(all_predictors())


trControl <- trainControl(method='cv', number = 10, savePredictions = TRUE, classProbs = TRUE)

# with the lda method, we can can specify the model just as we did with the glm function
lda_fit <- train(model_recipe, data = Train, method = model_type, trControl = trControl)

# the lda_fit object gives us some information about prior probabilities and coefficients of the linear discriminants
lda_fit$finalModel


# we can also use the predict function again to predict new outcomes
lda_class <- lda_fit %>% predict(type ='raw', newdata = Train) # this assumes a threshold of 0.5
lda_probs <- lda_fit %>% predict(type = 'prob', newdata = Train)

lda_probs[1:10, 2] # the '2' indicates the probability of predicting 'Yes'. 

# then, let's convert this to binary predictions with a default threshold value of 0.5
threshold <-  0.3517934 # from the threshold that maximizes the combined sensitivity and specificity
lda_pred <- factor(ifelse(lda_probs[, positive_class] > threshold, positive_class, negative_class) , levels = levels(Train[[target_var]]))


confusionMatrix(lda_pred, Train[[target_var]], positive = positive_class)

# ROC curve
lda_roc <- roc(Train[[target_var]], lda_probs[, positive_class], plot = TRUE, print.auc = TRUE, legacy.axes = TRUE, levels = c(negative_class, positive_class))
# get auc
(lda_Train_auc <- auc(lda_roc))
(threshold <- coords(lda_roc, 'best')$threshold)


#Prediction using the Test set

tlda_class <- lda_fit %>% predict(type ='raw', newdata = Test) # this assumes a threshold of 0.5
tlda_probs <- lda_fit %>% predict(type = 'prob', newdata = Test)

# show the first 10 predictions. 
tlda_probs[1:10, 2] # the '2' indicates the probability of predicting 'Yes'. 

# converting this to binary predictions with a default threshold value of 0.5
threshold <-  0.3347885 # from the threshold that maximizes the combined sensitivity and specificity
tlda_pred <- factor(ifelse(tlda_probs[, positive_class] > threshold, positive_class, negative_class) , levels = levels(Test[[target_var]]))

# Evaluation metrics confusion matrix from the model output:
postResample(tlda_pred, Test[[target_var]])

#Displaying the confusion matrix and additional metrics.
confusionMatrix(tlda_pred, Test[[target_var]], positive = positive_class)

# or  we can change that by manually providing the class
confusionMatrix(tlda_pred, Test[[target_var]], positive = "No")

# and looking at the ROC curve
(tlda_roc <- roc(Test[[target_var]], tlda_probs[, positive_class], plot = TRUE, print.auc = TRUE, legacy.axes = TRUE, levels = c(negative_class, positive_class)))

# getting the auc
(ldaTest_auc <- auc(tlda_roc))

# side note: how do we decide on the threshold to use?
# one option is: select the threshold that maximizes the combined sensitivity and specificity (other options available, see documentation)
(threshold <- coords(tlda_roc, 'best')$threshold)

```

## QDA

Let's look at an alternative to LDA: the Quadratic Discriminant Analysis. As the name implies, it uses non-linear (quadratic) decision boundaries instead of linear ones. We will again use the MASS library as it provides a qda() function for this.

```{r qda}

dataset <- NULL
census <- read_csv("Census.csv")
source("Utils.R") # load the functions provided in Utils.R
dataset <- prepare_census(census)


# Quadratic Discriminant Analysis
target_var <- 'Income'
model_form <- Income ~ Age + Sex + Race + MaritalStatus + WeeklyHours
model_type <- 'qda'
positive_class <- 'Yes'
negative_class <- 'No'

# QDA relies on full rank matrices, and if the underlying data (model) matrix does not have full rank, it will throw an error message. hence, let's catch this case using tryCatch so that we can still create the markdown document.
# Alternatively, and preferrably, one common thing to try is to change step_zv (remove zero variance) to 
# step_nzv (remove near zero variance), as near zero variance predictors often are not particularly useful but can create computation issues

model_recipe <- recipe(model_form, data = dataset) %>% 
  step_novel(all_predictors(), -all_numeric()) %>%
  step_unknown(all_predictors(), -all_numeric()) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>% 
  step_zv(all_predictors()) # or: step_nzv()

#Equal percentage split 
set.seed(3456)
train_index <- createDataPartition(y = dataset[[target_var]], p = 0.7, list = FALSE) # creata a 70-30 split

Train <- dataset %>% slice(train_index)
Test <- dataset %>% slice(-train_index)

tryCatch({
  
  trControl <- trainControl(method='cv', number = 10, savePredictions = TRUE, classProbs = TRUE)

  # with qhe lda method, we can can specify the model just as we did with the glm function
  qda_fit <- train(model_recipe, data = Train, method = model_type, trControl = trControl)
  
  # the qda_fit object gives us some information about prior probabilities and coefficients of the linear discriminants
  qda_fit$finalModel
  
  # we can also use the predict function again to predict new outcomes
  qda_class <- qda_fit %>%  predict(type ='raw', newdata = Train) # this assumes a threshold of 0.5
  qda_probs <- qda_fit %>% predict(type = 'prob', newdata = Train)
  
qda_probs[1:10, 2] # the '2' indicates the probability of predicting 'Yes'. 

threshold = 0.1656443
qda_pred <- factor(ifelse(qda_probs[, positive_class] > threshold, positive_class, negative_class), levels = levels(Train[[target_var]]))

confusionMatrix(qda_pred, Train[[target_var]], positive = positive_class)
  
  # ROC curve
  qda_roc <- roc(Train[[target_var]], qda_probs[, positive_class], plot = TRUE, print.auc = TRUE, legacy.axes = TRUE, levels = c(negative_class, positive_class))
  # get auc
  (qda_auc <- auc(qda_roc))
  (threshold <- coords(qda_roc, 'best')$threshold)

  
#Prediction using the Test set

tqda_class <- qda_fit %>% predict(type ='raw', newdata = Test) # this assumes a threshold of 0.5
tqda_probs <- qda_fit %>% predict(type = 'prob', newdata = Test)

# show the first 10 predictions. 
tqda_probs[1:10, 2] # the '2' indicates the probability of predicting 'Yes'. see the glm_probs dataset for further details

# then, let's convert this to binary predictions with a default threshold value of 0.5
threshold <-  0.1532303 # from the threshold that maximizes the combined sensitivity and specificity
tqda_pred <- factor(ifelse(tqda_probs[, positive_class] > threshold, positive_class, negative_class) , levels = levels(Test[[target_var]]))

# we can get the evaluation metrics confusion matrix from the model output:
postResample(tqda_pred, Test[[target_var]])

# we can use following function to display the confusion matrix and additional metrics.
confusionMatrix(tqda_pred, Test[[target_var]], positive = positive_class)

#  we can change that by manually providing the 'positive' class
confusionMatrix(tqda_pred, Test[[target_var]], positive = "No")

# and look at the ROC curve

(tqda_roc <- roc(Test[[target_var]], tqda_probs[, positive_class], plot = TRUE, print.auc = TRUE, legacy.axes = TRUE, levels = c(negative_class, positive_class)))
# get auc
(qdaTest_auc <- auc(tqda_roc))

# side note: how do we decide on the threshold to use?
# one option is: select the threshold that maximizes the combined sensitivity and specificity (other options available, see documentation)
(threshold <- coords(tqda_roc, 'best')$threshold)

}, error=function(cond){message(paste("error while creating the QDA model"))
})

```

## KNN (k-Nearest Neighbors)

Finally, let's see how we can implement knn classification in R. We'll use the knn function of the caret package for this.

```{r knn}


dataset <- NULL
census <- read_csv("Census.csv")
source("Utils.R") # load the functions provided in Utils.R
dataset <- prepare_census(census)


# KNN Analysis
target_var <- 'Income'
model_form <- Income ~ Age + Sex + Race + MaritalStatus + WeeklyHours
positive_class <- 'Yes'
negative_class <- 'No'
#threshold <- 0.5
model_type <- "knn"


#Equal percentage split 
set.seed(3456)
train_index <- createDataPartition(y = dataset[[target_var]], p = 0.7, list = FALSE) # creata a 70-30 split

Train <- dataset %>% slice(train_index)
Test <- dataset %>% slice(-train_index)


model_recipe <- recipe(model_form, data = dataset) %>%
  step_novel(all_predictors(), -all_numeric()) %>%
  step_unknown(all_predictors(), -all_numeric()) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>% 
  step_normalize(all_predictors()) %>%
  step_zv(all_predictors())

# as we need to evaluate different values of k, we can use the 'grid search' function of caret to loop through different values of k.

trControl <- trainControl(method='cv', number = 10, savePredictions = TRUE, classProbs = TRUE)

# let's build an initial model with k = 10
tuneGrid <- expand.grid(k = 10) # note: a grid search usually considers a set of parameters, not only one. However, in the current implementation of caret only one parameter is allowed when we don't use resampling
knn_fit <- train(model_recipe, data = Train, method = model_type, trControl = trControl, tuneGrid = tuneGrid)
knn_fit$finalModel 

# side note: if you want to automatically loop through multiple values of k, you need to delete the trControl parameter (courtesy of caret)
#tuneGrid <- expand.grid(k = seq(3,7,2)) 
#knn_fit <- train(model_recipe, data = dataset, method = model_type, tuneGrid = tuneGrid)
#knn_fit # note: the accuracy values shown here are much lower than the ones before, mostly because caret uses an implicit bootstrapping resampling to estimate the predicted accuracy

# get the probabilities and predictions

knn_fit_probs <- knn_fit %>% predict(type = "prob", newdata = Train) # due to a curiosity in caret, we need to specify the data on which we want to get the predictions. here: the entire dataset
knn_fit_class <- knn_fit %>% predict(type = 'raw', newdata = Train) 
knn_fit_probs[1:10, 2]


threshold = 0.2440476
knn_pred <- factor(ifelse(knn_fit_probs[, positive_class] > threshold, positive_class, negative_class), levels = levels(Train[[target_var]]))


confusionMatrix(knn_pred, Train[[target_var]], positive = positive_class)


# ROC curve
knn_roc <- roc(Train[[target_var]], knn_fit_probs[, positive_class], plot = TRUE, print.auc = TRUE, legacy.axes = TRUE, levels = c(negative_class, positive_class))
  # get auc
  (knn_auc <- auc(knn_roc))
#get best threshold
  (threshold <- coords(knn_roc, 'best')$threshold)

  
#Prediction using the Test set

tknn_class <- knn_fit %>% predict(type ='raw', newdata = Test) # this assumes a threshold of 0.5
tknn_probs <- knn_fit %>% predict(type = 'prob', newdata = Test)

# show the first 10 predictions. 
tknn_probs[1:10, 2] # the '2' indicates the probability of predicting 'Yes'. see the glm_probs dataset for further details

# then, let's convert this to binary predictions with a default threshold value of 0.5
threshold <-  0.2264957 # from the threshold that maximizes the combined sensitivity and specificity
tknn_pred <- factor(ifelse(tknn_probs[, positive_class] > threshold, positive_class, negative_class) , levels = levels(Test[[target_var]]))

# we can get the evaluation metrics confusion matrix from the model output:
postResample(tknn_pred, Test[[target_var]])

# we can use following function to display the confusion matrix and additional metrics.
confusionMatrix(tknn_pred, Test[[target_var]], positive = positive_class)

#  we can change that by manually providing the 'positive' class
confusionMatrix(tknn_pred, Test[[target_var]], positive = "No")

# and look at the ROC curve

(tknn_roc <- roc(Test[[target_var]], tknn_probs[, positive_class], plot = TRUE, print.auc = TRUE, legacy.axes = TRUE, levels = c(negative_class, positive_class)))
# get auc
(knnTest_auc <- auc(tknn_roc))

# side note: how do we decide on the threshold to use?
# one option is: select the threshold that maximizes the combined sensitivity and specificity (other options available, see documentation)
(threshold <- coords(tknn_roc, 'best')$threshold)

```

## The end

